---
title: "EDA project"
author: "Abhinav Ache"
date: '2022-11-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(gapminder)
library(GGally)
library(MASS)
```

```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
Happiness_data = read_xls('Happiness_data.xls')  #data from https://worldhappiness.report/
europe_data = read_xls('sub_europe.xls') 
```

```{r}
names(Happiness_data) <- c("country", "year", "happiness_score", "Log_GDP_per_capita", "social_support", "life_expectancy", "freedom_to_make_life_choices", "generosity", "perceptions_of_corruption", "positive_affect", "negative_affect", "confidence_in_national_government")
```

```{r}
# Extracting only the required columns
Happiness_new = Happiness_data[, c("country", "year", "happiness_score", "Log_GDP_per_capita", "life_expectancy")]
```

```{r}
# Removing the null values
Happiness_new = na.omit(Happiness_new)
```

```{r}
# Pairwise Plots
ggpairs(Happiness_new[, c("happiness_score", "Log_GDP_per_capita", "life_expectancy")])+
  ggtitle("Fig 1: Pairwise plots between 'Happiness score','GDP per capita' and 'life expectancy'")
```

# Distribution of the data

```{r}
summary(Happiness_new)
```


```{r}
hist(Happiness_new$year, xlab = "year Score", main = "Distribution of year")
```

```{r}
hist(Happiness_new$happiness_score, xlab = "Hapiness Score", main = "Distribution of Happiness score")
```

```{r}
hist(Happiness_new$Log_GDP_per_capita, xlab = "GDP per capita(log scale)", main = "Distribution of GDP per capita")
```

```{r}
hist(Happiness_new$life_expectancy, xlab = "life expectancy", main = "Distribution of life expectancy")
```




```{r}
Happiness_new = Happiness_new %>% filter(Log_GDP_per_capita >= 6)
```


# Finding the relationship between Log_GDP and Happiness Score



```{r}
ggplot(Happiness_new, aes(x = Log_GDP_per_capita, y = happiness_score)) + geom_point(alpha = 0.4,size = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  xlab('GDP per capita, log scale') +
  scale_color_manual(values = cbPalette)+
  ylab('Happiness Score') +
  ggtitle('Figure 2: Log GDP per capita vs. Happiness score') + 
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Orange line indicates a linear model, while Blue curve indicates a loess fit")
```

```{r}
ggplot(Happiness_new, aes(x = Log_GDP_per_capita, y = happiness_score)) + geom_point(alpha = 0.3,size = 0.6) +
  facet_wrap(~cut_number(year,n=4),ncol=2) + geom_smooth(method = "lm", se = FALSE,size =0.85) +
  geom_smooth(method = "gam",se = FALSE , span =0.5,color= "orange",size =0.85)+
  geom_smooth(method="loess",se = FALSE,color="green",size =0.85)+
  xlab('GDP per capita, log scale') +
  ylab('Happiness Score') +
  ggtitle('Figure 3: Log GDP per capita vs. Happiness score for different time periods') +
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Blue line indicates a linear model,Orange Indicates GAM, Green indicates loess")
```

```{r}
Happiness_new = Happiness_new %>% filter(year != 2005)
```


# Adding continent data for better analysis 

Adding Continent column from the gapminder dataset. 
```{r}
continents = gapminder[, c("country", "continent")]
```



```{r}
continents = continents[!duplicated(continents$country), ]
```

```{r}
df = merge(x = Happiness_new, y = continents, by = "country")
```

Check for null values
```{r}
sum(is.na(df[5]))
```



# How does happiness score vary with GDP per capita across various continents?

```{r}
df %>% filter(continent != "Oceania") %>%
        ggplot(aes(x = Log_GDP_per_capita, y = happiness_score)) + geom_point(alpha = 0.5,size = 0.8) +
          facet_wrap(~continent) +
  geom_smooth(se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  xlab('GDP per capita, log scale') +
  ylab('Happiness Score') +
  ggtitle('Figure 4: Log GDP per capita vs. Happiness score faceted by continent') +
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Orange line indicates a linear model, while Blue curve indicates a loess fit")
```


# Dividing the europe into the sub regions and checking the relationship bwtween happiness_score and Log_GDP_per_Capita

```{r}
europe = filter(df,continent=='Europe')
#europe
```


```{r}
europe_subregion = europe_data[!duplicated(europe_data$country), ]
#europe_subregion
```

```{r}
europe_new = merge(x = europe, y = europe_subregion, by = "country")
```



```{r}
ggplot(europe_new,aes(x = Log_GDP_per_capita, y = happiness_score)) + geom_point() +
          facet_wrap(~region) +
  geom_smooth(se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  xlab('GDP per capita, log scale') +
  ylab('Happiness Score')
```

# Dividing western_europe and Northern Europe based on different time periods and checking the relationship


```{r}
europe_new %>% filter(region == "Northern Europe")%>%
  ggplot(aes(x=Log_GDP_per_capita,y=happiness_score))+geom_point()+geom_smooth()+
  facet_grid(~cut_number(year,n=3))+
  geom_smooth(method="lm",color="orange")
  
```


```{r}
europe_new %>% filter(region == "Western Europe")%>%
  ggplot(aes(x=Log_GDP_per_capita,y=happiness_score))+geom_point()+geom_smooth(se = FALSE)+
  facet_grid(~cut_number(year,n=3))+
  geom_smooth(method="lm",color="orange",se= FALSE)
  
```


## Relationship between Happiness_Score and Life_expectancy

```{r}
df %>% filter(continent != "Oceania")%>%
  filter(country != "Haiti") %>%
  ggplot(aes(x =life_expectancy, y = happiness_score)) + geom_point(size = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  geom_smooth(method="gam",se=FALSE,color = "green")+
  #geom_smooth(method="rlm",se=FALSE,color="violet")+
  xlab('Life Expectancy') +
  ylab('Happiness Score') +
  ggtitle('Figure 6: Life Expectancy vs. Happiness score') + 
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Orange line indicates a linear model, while Blue curve indicates a loess fit,green = GAM, Violet = rlm")
```

lets facet by year and see

```{r}
df %>% filter(continent != "Oceania")%>%
  filter(country != "Haiti") %>%ggplot( aes(x =life_expectancy, y = happiness_score)) + geom_point(size = 0.8) +
  facet_grid(~cut_number(year,n=4))+
  geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  geom_smooth(method="gam",se=FALSE,color = "green")+
  #geom_smooth(method="rlm",se=FALSE,color="violet")+
  xlab('Life_Expectancy') +
  ylab('Happiness Score') +
  ggtitle('Fig 7: Life expectancy vs. Happiness score facted By different time periods') + 
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Orange line indicates a linear model, while Blue curve indicates a loess fit,green indicates GAM fit")


```






```{r}
ggplot(df, aes(x = Log_GDP_per_capita, y = happiness_score)) + geom_point() +
  facet_wrap(~cut_number(life_expectancy,n=6),ncol=2)+
  geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  geom_smooth(method="gam",se=FALSE,color = "green")+
  geom_smooth(method="rlm",se=FALSE,color="violet")+
  xlab('Log_GDP') +
  ylab('Happiness Score') +
  ggtitle('Log GDP per capita vs. Happiness score conditioned on life_expectancy') + 
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Orange line indicates a linear model, while Blue curve indicates a loess fit")

```



```{r}
df %>% filter(continent != "Oceania")%>%
  filter(country != "Haiti") %>%
  ggplot(aes(x = life_expectancy, y = happiness_score)) + geom_point(alpha = 0.5,size = 0.9) +
  facet_wrap(~continent,ncol=2)+
  #facet_grid(~cut_number(life_expectancy,n=6))+
  geom_smooth(method = "loess", se = FALSE,color="orange") +
  geom_smooth(method = "lm", se = FALSE) +
  #geom_smooth(method="gam",se=FALSE,color = "green")+
  #geom_smooth(method="rlm",se=FALSE,color="violet")+
  xlab('life_expectancy') +
  ylab('Happiness Score') +
  ggtitle('Figure 8: Life expectancy vs. Happiness score faceted by continents') + 
  labs(caption = "Data Source: https://worldhappiness.report/", subtitle = "Orange line indicates a linear model, while Blue curve indicates a loess fit")
  
```
```{r}
df2 = df %>% filter(continent != "Oceania")%>%
  filter(country != "Haiti")
```



# Creating a models

```{r}
model1 = lm(happiness_score ~ Log_GDP_per_capita+life_expectancy+continent+Log_GDP_per_capita:continent+life_expectancy:continent, data = df2)
```


```{r}
library(mgcv)
```

```{r}
model2 = gam(happiness_score ~ Log_GDP_per_capita+s(life_expectancy)+continent:Log_GDP_per_capita, data = df2)
```

```{r}
plot(model2)
```

```{r}
summary(model1)
```


```{r}
summary(model2)
```

# Plotting the co-efficients for model 1

```{r}
model1.effects = dummy.coef(model1)
model1.effects
```



```{r}
model1.effects$`Log_GDP_per_capita:continent`+model1.effects$Log_GDP_per_capita
```

```{r}
model1_gdg_slopes = as.vector(model1.effects$`Log_GDP_per_capita:continent`+model1.effects$Log_GDP_per_capita)
model1_life = as.vector(model1.effects$`life_expectancy:continent` +model1.effects$life_expectancy)
continent = c("Africa","Americas","Asia","Europe")
model1.coeff = data.frame(continent,Log_GDP_per_capita = model1_gdg_slopes, life_expectancy = model1_life )
```

```{r}
ggplot(model1.coeff,aes(x=continent,y=Log_GDP_per_capita))+geom_point()+ylab("Co-efficients for 'Log GDP per capita'")+
  ggtitle("Model 1: Co-efficients of 'Log GDP per capita' for different continents")
```

```{r}
ggplot(model1.coeff,aes(x=continent,y=life_expectancy))+geom_point()+ylab("Co-efficients for 'Life expectancy'")+
  ggtitle(" Model 1: Co-efficients of 'Life Expectancy' for different continents")
```

# Plotting the co-efficients for model 2

```{r}
model2.effects = dummy.coef(model2)
model2.effects
```


```{r}
model2_gdg_slopes = as.vector(model2.effects$`Log_GDP_per_capita:continent`+model2.effects$Log_GDP_per_capita)
continent = c("Africa","Americas","Asia","Europe")
model2.coeff = data.frame(continent,Log_GDP_per_capita = model2_gdg_slopes )
```


```{r}
ggplot(model2.coeff,aes(x=continent,y=Log_GDP_per_capita))+geom_point()+ylab("Co-efficients for 'Log GDP per capita'")+
  ggtitle(" Co-efficients of 'Log GDP per capita' for different continents")
```

## Visualization of Fit for Model 1

```{r}
data.grid = expand.grid(continent = unique(df2$continent),
                        Log_GDP_per_capita = seq(6,12,1),
                        life_expectancy = seq(45,80,5))
```

```{r}
data.predict = predict(model1, newdata = data.grid)

data.grid$.fit = as.vector(data.predict)

```

```{r}
# As all the countries in Europe have GDP higher than 9 in all the years, I want to remove countries in Europe with GDP lower than 9 since they may give incorrect or negative predictions
data.grid %>% filter( .fit>2)%>% 
  ggplot(aes(x=Log_GDP_per_capita,y= .fit,color = continent))+geom_line()+facet_wrap(~cut_number(life_expectancy,n=8),ncol=4)
```

```{r}
# As all the countries in Europe have GDP higher than 9 in all the years, I want to remove countries in Europe with GDP lower than 9 since they may give incorrect or negative predictions
data.grid %>% filter( .fit>2)%>% 
  ggplot(aes(x=Log_GDP_per_capita,y= .fit,group = life_expectancy,color = life_expectancy))+
  geom_line()+facet_wrap(~continent,ncol=2)+scale_color_viridis_c()+ylab("fitted Values")+
  ggtitle("Fig 9: Visualization of fit for model 1")
```

## Visualization of Fit for Model 2
```{r}
data.grid2 = expand.grid(continent = unique(df2$continent),
                        Log_GDP_per_capita = seq(6,12,1),
                        life_expectancy = seq(45,80,5))
```


```{r}
data.predict2 = predict(model2, newdata = data.grid2)

data.grid2$.fit = as.vector(data.predict2)

```


```{r}
# As all the countries in Europe have GDP higher than 9 in all the years, I want to remove countries in Europe with GDP lower than 9 since they may give incorrect or negative predictions
data.grid2 %>% filter( .fit>2)%>% 
  ggplot(aes(x=life_expectancy ,y= .fit,group = Log_GDP_per_capita,color = Log_GDP_per_capita))+
  geom_line()+facet_wrap(~continent)+
  scale_color_viridis_c()+
  ggtitle("Fig 10: Visualization of fit for model 2")
```


# seeing the fitted values of all the above models for year 2018


```{r}
model1.df = data.frame(df2,.fitted = fitted.values(model1),.resid =residuals(model1))
```
```{r}
model2.df = data.frame(df2,.fitted = fitted.values(model2),.resid =residuals(model2))

```


## plotting Fitted Values vs happiness score 

```{r}
model1.df %>% filter(year==2018)%>%
  ggplot(aes(y=.fitted,x=happiness_score))+geom_point()
```

```{r}
model2.df %>% filter(year==2018)%>%
  ggplot(aes(y=.fitted,x=happiness_score))+geom_point()
```


## plotting Fitted Values vs Log_GDP


```{r}
model1.df %>% filter(year==2018)%>%
  ggplot(aes(y=.fitted,x=Log_GDP_per_capita,color=continent))+geom_point()+scale_color_manual(values = cbPalette)+geom_smooth()+ggtitle("Fitted Values for year 2018 with Model 1")
```

```{r}
model2.df %>% filter(year==2018)%>%
  ggplot(aes(y=.fitted,x=Log_GDP_per_capita,color=continent))+geom_point()+scale_color_manual(values = cbPalette)+geom_smooth(se =FALSE)
```

## checking the residuals for different models


```{r}
model1.df %>% filter(year==2018)%>%
  ggplot(aes(y=.resid,x=Log_GDP_per_capita))+geom_point()+scale_color_manual(values = cbPalette)+geom_hline(yintercept = 0)+geom_smooth(se = FALSE)
```


```{r}
model2.df %>% filter(year==2018)%>%
  ggplot(aes(y=.resid,x=Log_GDP_per_capita))+geom_point()+scale_color_manual(values = cbPalette)+geom_hline(yintercept = 0)+geom_smooth(se = FALSE)
```

## checking Fitted vs Residuals

```{r}
model1.df %>% filter(year==2018)%>%
  ggplot(aes(y=.resid,x= .fitted))+geom_point()+scale_color_manual(values = cbPalette)+geom_smooth(method = "lm",se = FALSE)
```


```{r}
model2.df %>% filter(year==2018)%>%
  ggplot(aes(y=.resid,x= .fitted))+geom_point()+scale_color_manual(values = cbPalette)+geom_smooth(method = "lm",se = FALSE)
```

